Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-11 manuscript/main.tex
complete discussion

Among these, \emph{A. hadrus} is the variable with 
highest average relative importance along all reported 
(mean = 1.4\%; min = 0.8\%; max = 3.6\%, Figure 4D-E).
\emph{B. wexlerae}, which was important in RF models of cognitive function, 
was also important in models of the entorhinal, fusiform and lingual areas
and had the highest importance in models predicting the size of
both the left and right right parahippocampus.
Another cluster of model importance contained taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains \emph{B. ovatus} and \emph{B. uniformis},
but also \emph{Alistipes finegoldii and S. salivarius}.

@@

We did not identify any statistical associations between individual species
and cognitive performance in our youngest subset (0 to 6 months).
We hypothesize that this may be because brain developmental trajectories
begin \textit{in utero}, and microbial influences (that can only begin at birth)
take additional time to manifest.
In line with this hypothesis, RF models that were able to learn associations
between early-life microbiomes and future cognitive function scores (Figure 3).
The top-ranked species for these models included several 
taxa that were important for predicting concurrent cognitive assessment scores after 18 months, notably
pathobiontic \textit{Enterobacteriaceae} such as
\textit{Escherichia} and \textit{Klebsiella}. 
Representatives of these genera are known to be early and long-term persistent
colonizers of infant intestine that can, in some settings,
promote inflammatory response \cite{popeEnterobacEarlyGut2019}.
By contrast, several inflammation-reducing representatives
of the \textit{Bifidobacterium} genus,
from the early HMO-metabolizer \textit{B. longum} to the
later complex carbohydrate relying \textit{B. pseudocatenulatum} \cite{chungBifidoPsudocatenulatum2021},
commonly associated with weaning to solid food in early age,
also showed high relative importance.
The involvement of most important taxa in different aspects of inflammatory response
resonate with previous findings that associate chronic inflammation with
changes to neural responses in some types of stimulae, notably visual\cite{xieChronicInfNeuro2019}.

@@

Our ability to probe individual cognitive subscales provided
results that also align with previous findings and provide insight to
potential gut-microbial connections between cognition and neuroanatomy.
For example, a longitudinal neurocognitive development study pointed at bilateral
associations between the thalamus and the expressive language
component of the Mullen scales \cite{chenLongitudinalBrainMullen2023}.
In line with this finding, we found that taxa such as 
\textit{B. wexlerae} and \textit{A. hadrus} to be among the top 15\% important predictors
on both left and right thalamus and the EL subscale.
Another example of those findings is a previously reported association between
the central opercular cortex (COC) and gross motor development \cite{jiangBrainNetworkCognition2023}. In this case, we identified that
butyrate-producing \textit{Fusicatenibacter saccharivorans} is among the 15\% most important predictors
for both the gross motor subscale and three bilateral components of the COC
(left and right pars orbitalis, pars opercularis and pars triangularis).
Previous results also link to relationships between the caudate and the cerebellum,
and the Visual Reception score \cite{deoniMyelinationMullenPreprint2022}.
\textit{S. salivarius} was consistently found among the top 20\% predictors for this subscale, along with
both sides of the thalamus and cerebellum white matter, along with cerebellar vermal lobules I-X.

>>>>

Among these, \emph{A. hadrus} is the variable with 
highest average relative importance along all reported models
(mean = 1.4\%; min = 0.8\%; max = 3.6\%, Figure 4D-E).
\emph{B. wexlerae}, which was important in RF models of cognitive function, 
was also important in models of the entorhinal, fusiform and lingual areas
and had the highest importance in models predicting the size of
both the left and right right parahippocampus.
Another cluster of model importance contained taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains \emph{B. ovatus} and \emph{B. uniformis},
but also \emph{Alistipes finegoldii and S. salivarius}.

@@

We did not identify any statistical associations between individual species
and cognitive performance in our youngest subset (0 to 6 months).
We hypothesize that this may be because brain developmental trajectories
begin \textit{in utero}, and microbial influences (that can only begin at birth)
take additional time to manifest.
In line with this hypothesis, RF models that were able to learn associations
between early-life microbiomes and future cognitive function scores (Figure 3).
The top-ranked species for these models included several 
taxa that were important for predicting concurrent cognitive assessment scores after 18 months, notably
pathobiontic \textit{Enterobacteriaceae} such as
\textit{Escherichia} and \textit{Klebsiella}. 
Representatives of these genera are known to be early and long-term persistent
colonizers of infant intestine that can, in some settings,
promote inflammatory response \cite{popeEnterobacEarlyGut2019}.
By contrast, several inflammation-reducing representatives
of the \textit{Bifidobacterium} genus,
from the early HMO-metabolizer \textit{B. longum} to the
later complex carbohydrate relying \textit{B. pseudocatenulatum} \cite{chungBifidoPsudocatenulatum2021},
commonly associated with weaning to solid food in early age,
also showed high relative importance.
The involvement of most important taxa in different aspects of inflammatory response
resonate with previous findings that associate chronic inflammation with
changes to neural responses in some types of stimulae \cite{xieChronicInfNeuro2019}.

@@

Our ability to probe individual cognitive subscales provided
results that also align with previous findings and provide insight to
potential gut-microbial connections between cognition and neuroanatomy.
For example, a longitudinal neurocognitive development study pointed at bilateral
associations between the thalamus and the expressive language
component of the Mullen scales \cite{chenLongitudinalBrainMullen2023}.
In line with this finding, we found that taxa such as 
\textit{B. wexlerae} and \textit{A. hadrus} to be among the top 15\% important predictors
on both left and right thalamus and the expressive language subscale.
Another example of those findings is a previously reported association between
the central opercular cortex (COC) and gross motor development \cite{jiangBrainNetworkCognition2023}.
In this case, we identified that
butyrate-producing \textit{F. saccharivorans} is among the 15\% most important predictors
for both the gross motor subscale and three bilateral components of the COC
(left and right pars orbitalis, pars opercularis and pars triangularis).
Previous results also link to relationships between the caudate and the cerebellum,
and the visual reception score \cite{deoniMyelinationMullenPreprint2022}.
\textit{S. salivarius} was consistently found among the top 20\% predictors for this subscale, along with
both sides of the thalamus and cerebellum white matter, along with cerebellar vermal lobules I-X.
